---
title: "Batch-aware QC for multi-sample scRNA-seq with scBatchQC"
author:
  - name: "Subhadip Jana"
    email: "subhadipjana1409@gmail.com"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Batch-aware QC for multi-sample scRNA-seq with scBatchQC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 7,
    fig.height = 5,
    message = FALSE,
    warning = FALSE
)
```

# Introduction

In multi-batch single-cell RNA-sequencing experiments, cell quality
control is often performed independently per sample before data
integration. This approach has a well-known but rarely addressed flaw:
standard MAD-based outlier thresholds (e.g. `scuttle::isOutlier`) are
estimated from the global distribution and do not account for
systematic differences in QC metric distributions across batches.

Consider a study with two batches:

- **Batch A**: high-depth 10x v3 libraries from fresh tissue.
- **Batch B**: lower-depth 10x v2 libraries from cryopreserved samples.

Applying a single global library-size threshold will either
over-filter Batch B (removing real cells) or under-filter Batch A
(keeping low-quality cells) — neither is desirable.

`scBatchQC` addresses this by:

1. Computing per-batch QC metric location (median) and scale (MAD).
2. Shrinking per-batch estimates toward a global empirical Bayes prior,
   borrowing strength across batches.
3. Estimating per-batch doublet rates as a function of cells loaded and
   protocol type.
4. Returning a `SingleCellExperiment` with calibrated QC flags in
   `colData`.

# Installation

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
BiocManager::install("scBatchQC")
```

# Quick start

```{r quickstart}
library(scBatchQC)
library(SingleCellExperiment)

# Simulate a two-batch dataset
set.seed(123)
n_genes <- 500
n_cells <- 200

# Batch 1: higher depth
counts_b1 <- matrix(rpois(n_genes * 100, lambda = 12),
    nrow = n_genes, ncol = 100
)
# Batch 2: lower depth + a few low-quality cells
counts_b2 <- matrix(rpois(n_genes * 100, lambda = 5),
    nrow = n_genes, ncol = 100
)
counts_b2[, 1:5] <- counts_b2[, 1:5] / 10 # inject low-quality cells

counts <- cbind(counts_b1, counts_b2)
rownames(counts) <- paste0("Gene", seq_len(n_genes))
rownames(counts)[1:30] <- paste0("MT-", seq_len(30))
colnames(counts) <- paste0("Cell", seq_len(n_cells))

sce <- SingleCellExperiment(assays = list(counts = counts))
sce$batch <- rep(c("Batch1_v3", "Batch2_v2"), each = 100)
sce
```

# Batch-aware QC metrics

The central function `batchAwareQCMetrics()` computes per-cell QC
metrics and flags outliers using batch-harmonized MAD thresholds.

```{r qc_metrics}
sce <- batchAwareQCMetrics(
    sce,
    batch           = "batch",
    nmads           = 3,
    shrink_strength = 0.5
)

# QC columns added to colData
grep("^scBatchQC", names(colData(sce)), value = TRUE)
```

```{r qc_table}
# Outlier summary per batch
table(sce$scBatchQC_outlier, sce$batch)
```

## How shrinkage works

The `shrink_strength` parameter controls the balance between
per-batch estimation and global pooling:

- `shrink_strength = 0`: fully per-batch (equivalent to running
  `scuttle::isOutlier` per batch separately).
- `shrink_strength = 1`: full pooling (equivalent to global MAD,
  ignoring batch).
- `shrink_strength = 0.5` (default): empirical Bayes midpoint.

For batches with very few cells (< 50), we recommend increasing
`shrink_strength` to 0.7–0.9.

# Visualizing QC distributions

```{r plot_qc, fig.cap = "QC metric distributions per batch. Red dashed lines indicate batch-harmonized upper thresholds. Red points are cells flagged as outliers."}
plotBatchQC(sce, batch = "batch")
```

# Estimating batch doublet rates

```{r doublet_rates}
cells_loaded <- c(Batch1_v3 = 8000, Batch2_v2 = 5000)

sce <- estimateBatchDoubletRate(
    sce,
    batch        = "batch",
    cells_loaded = cells_loaded,
    protocol     = c(Batch1_v3 = "10x_v3", Batch2_v2 = "10x_v2")
)

# Per-batch estimated doublet rates (as proportion of recovered cells)
tapply(sce$scBatchQC_doublet_rate, sce$batch, unique)
```

The doublet rate model follows the empirical relationship from 10x
Genomics documentation: approximately 0.8% doublets per 1,000 cells
loaded on the Chromium controller.

# Exploring thresholds interactively

`harmonizeQCThresholds()` lets you explore how the number of flagged
cells changes across different MAD multipliers without re-running the
full pipeline.

```{r threshold_sweep}
# Sweep from 2 to 4 MADs
results <- lapply(c(2, 2.5, 3, 3.5, 4), function(n) {
    r <- harmonizeQCThresholds(sce, batch = "batch", nmads = n)
    data.frame(
        nmads = n,
        total_flagged = sum(rowSums(as.data.frame(r$n_flagged)))
    )
})
do.call(rbind, results)
```

# Recommended workflow

The recommended full workflow for a multi-batch scRNA-seq study is:

```{r full_workflow, eval = FALSE}
# Step 1: batch-aware QC flagging
sce <- batchAwareQCMetrics(sce, batch = "batch", nmads = 3)

# Step 2: estimate doublet rates
sce <- estimateBatchDoubletRate(sce,
    batch = "batch",
    cells_loaded = my_cells_loaded
)

# Step 3: visualise and inspect
plotBatchQC(sce, batch = "batch")

# Step 4: remove flagged cells
sce_filtered <- sce[, !sce$scBatchQC_outlier]

# Step 5: downstream analysis (normalisation, integration, etc.)
# -> pass sce_filtered to scran, Seurat, Harmony, etc.
```

# Session information

```{r session}
sessionInfo()
```
